---
title: "paper_replication"
output: pdf_document
---
REPLICATION:
TODO: averages of averages vs. averages for table 2 
TODO: possibly email Alice asking about this county vs. state table 2 issue <- Jo 
TODO: write up <- Toby
TODO: find abortion data used by Brown et al. to compare in discussion <- Toby

EXTENSION:
TODO: use all counties we have data on instead of just those 18 states
TODO: rescore the high vs. low restriction variable to make it make more sense
TODO: add in rural vs urban, % college grad, more variables do a quick variable selection
TODO: repeat for mortality and fertility <- Jo look for this data 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#load packages
options(tidyverse.quiet=TRUE)
library(tidyverse)
library(readxl)
library(dplyr)
library(geosphere)
library(stringr)
library(kableExtra)
library(table1)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
#load in relevant data
providers <- read.csv("provider_list_geo.csv")
countycentroids <- read.csv("county_pop_centroids.csv")
countydem <- read.csv("county_demographics.csv")
abortions <- read.csv("cleanedabortiondata.csv")
restrictions <- read.csv("state_restrictions2010.csv")
countypres <- read.csv("countypres.csv")
```


```{r}
########### clean county dem data ###########
countydem <- read.csv("county_demographics.csv")

for (i in 1:nrow(countydem)) {
  countydem$state[i] <- unlist(strsplit(countydem$NAME[i], ", "))[length(unlist(strsplit(countydem$NAME[i], ", ")))]
}

countydem$county <- gsub(countydem$NAME, pattern = ", [[:print:]]*$", replacement = "")
countydem$county <- gsub(countydem$county, pattern = " County", replacement = "")

countydem <- countydem %>% filter(countydem$state %in% state.name)

for (i in 1:nrow(countydem)) {
  countydem$state.abb[i] <- state.abb[which(state.name == countydem$state[i])]
}

countydem$county <- str_to_title(countydem$county)

```


```{r}
########### calculate abortion rates in 2010 ###########

#filter abortion rate data to only 2010
abortions2010 <- abortions %>% dplyr::select(state, county, X2010)

length(unique(abortions2010$state)) #only 42 states in abortion data

#look at percent missing abortion counts by state
aggregate(X2010 ~ state, data = abortions2010, function(x) {sum(!(is.na(x)))/length(x)}, na.action = NULL)

# clean abortion data to make counties and states match county dem data
abortions2010$county <- str_to_title(abortions2010$county)
abortions2010$county <- gsub(abortions2010$county, pattern = "[()]", replacement = "")
abortions2010$county <- gsub(abortions2010$county, pattern = "Ind. City", replacement = "City")
abortions2010$county <- gsub(abortions2010$county, pattern = "Saint", replacement = "St.")
abortions2010$county <- gsub(abortions2010$county, pattern = "Miami Dade", replacement = "Miami-Dade")
abortions2010$county <- gsub(abortions2010$county, pattern = "Hawai'i", replacement = "Hawaii")
abortions2010$county <- gsub(abortions2010$county, pattern = "Kaua'i", replacement = "Kauai")
abortions2010$county <- gsub(abortions2010$county, pattern = "St.e Genevieve", replacement = "Ste. Genevieve")
abortions2010$county <- gsub(abortions2010$county, pattern = "Shannon/Oglala Lakota", replacement = "Oglala Lakota")

#fix NA values and make abortion counts numeric
abortions2010$X2010 <- na_if(abortions2010$X2010, "?")
abortions2010$X2010 <- na_if(abortions2010$X2010, "")
abortions2010$X2010 <- as.numeric(abortions2010$X2010)

#remove non-counties
abortions2010 <- abortions2010 %>% filter(!(county %in% c("Nonresidents", "Protected", "Other States", "Other Countries", "Non-Residents", "Other", "Nevada Protected", "Nevada Unspecified", "Nebraska Protected", "County", "Maryland")))

#look at counties in abortion data but not countydem data
abortions2010$county[!(abortions2010$county %in% countydem$county)]
#"St. Louis County", "Bedford City", "Clifton Forge City"

#look at counties in countydem data but not abortion data
countydem$county[!(countydem$county %in% abortions2010$county)] #many 
unique(countydem$state.abb[!(countydem$state.abb %in% abortions2010$state)]) #"AK" "CT" "IA" "LA" "MA" "NH" "NM" "WY"

#Add abortion counts to county data
for (i in 1:nrow(countydem)) {
  if (length(abortions2010$X2010[abortions2010$state == countydem$state.abb[i] & abortions2010$county == countydem$county[i]]) == 0) {
    countydem$AbortionCount2010[i] <- NA
  }
  else {
    countydem$AbortionCount2010[i] <- abortions2010$X2010[abortions2010$state == countydem$state.abb[i] & abortions2010$county == countydem$county[i]]
  }
}

#calculate abortion rates
countydem <- countydem %>% mutate(AbortionRate2010 = AbortionCount2010/Women)

#look at percent missing abortion rates by state
aggregate(AbortionRate2010 ~ state, data = countydem, function(x) {sum(!(is.na(x)))/length(x)}, na.action = NULL)

```

```{r}
########### calculate percent democratic votes by county in in 2008 ###########
countypresdem2008 <- countypres %>% filter(party == "DEMOCRAT" & year == 2008)
countypresdem2008$county_name <- str_to_title(countypresdem2008$county_name)

countydem$county <- gsub(countydem$county, pattern = " Parish", replacement = "")
countydem$county <- gsub(countydem$county, pattern = " City", replacement = "")
countypresdem2008$county_name <- gsub(countypresdem2008$county_name, pattern = " City", replacement = "")
countypresdem2008$county_name <- gsub(countypresdem2008$county_name, pattern = "Saint", replacement = "St.")
countypresdem2008$county_name <- gsub(countypresdem2008$county_name, pattern = " County", replacement = "")

missingcounties <- countypresdem2008[!(countypresdem2008$county_name %in% countydem$county), ] #all of AK and a few others (DC, Dona Ana New Mexico, Kansas City Missouri, Maine Uocava Maine, Federal Precinct Rhode Island)

countypresdem2008 <- countypresdem2008 %>% mutate(percentdem2008 = candidatevotes/totalvotes) 
countypresdem2008 <- countypresdem2008 %>% rename(state.abb = state_po, county = county_name) %>% dplyr::select(state.abb, county, percentdem2008)

countydem <- left_join(countydem, countypresdem2008, by = c("state.abb", "county"))

#look at percent missing dem vote by state
aggregate(percentdem2008 ~ state, data = countydem, function(x) {sum(!(is.na(x)))/length(x)}, na.action = NULL)

```

```{r}
########### calculate distances from county centroids to closest provider location ###########

for (i in 1:nrow(countycentroids)) {
  lat <- countycentroids$LATITUDE[i]
  long <- countycentroids$LONGITUDE[i]
  dists <- distHaversine(cbind(long, lat), cbind(providers$Lon, providers$Lat))
  countycentroids$distancetoprovider[i] <- min(dists)
}

countycentroids <- countycentroids %>% rename(county = COUNAME, state = STNAME)
countycentroids$county <- str_to_title(countycentroids$county)
countycentroids <- countycentroids %>% dplyr::select(county, state, distancetoprovider)
countycentroids <- countycentroids %>% filter(state != "Puerto Rico")
countycentroids$county <- gsub(countycentroids$county, pattern = " City", replacement = "")

missingcounties <- countycentroids[!(countycentroids$county %in% countydem$county), ] #Alaska and DC

countydem <- left_join(countydem, countycentroids, by = c("state", "county"))

#look at percent missing distance by state
aggregate(distancetoprovider ~ state, data = countydem, function(x) {sum(!(is.na(x)))/length(x)}, na.action = NULL)

```

```{r}
#make new variable for high vs. low restrictions
restrictions <- restrictions %>% mutate(totalrestrict = parental_involvement + two_trips + trap + unconstitutional_ban, 
                                        restrictionlevel = case_when(totalrestrict >= 3 ~ "high", 
                                                              totalrestrict < 3 ~ "low"))
restrictions$state[!(restrictions$state %in% countydem$state)] #none
countydem <- left_join(countydem, restrictions, by = "state")
countydem$restrictionlevel <- factor(countydem$restrictionlevel, levels = c("low", "high"))

```

```{r}
#look at demographics

countydem <- countydem %>% mutate(inanalysis = case_when(state.abb %in% c("AZ", "DE", "GA", "IL", "IN", "MI", "NY", "NC", "OH", "OK", "OR", "PA", "SC", "TX", "UT", "VT", "WA", "WI") ~ 1, !(state.abb %in% c("AZ", "DE", "GA", "IL", "IN", "MI", "NY", "NC", "OH", "OK", "OR", "PA", "SC", "TX", "UT", "VT", "WA", "WI")) ~ 0))

countydemanalysis <- countydem %>% filter(inanalysis == 1)

summarytable <- countydemanalysis %>% group_by(state) %>% summarize(totalcounties = n())
usedcounties <- aggregate(AbortionRate2010 ~ state, data = countydemanalysis, function(x) {sum(!(is.na(x)))}, na.action = NULL)
summarytable <- left_join(summarytable, usedcounties, by = "state")
summarytable %>% mutate(percent = AbortionRate2010/totalcounties) %>% kbl(col.names = c("State", "Total Counties", "Counties with Abortion Data in 2010", "Percent"), caption = "Table 1. Number of Counties with Abortion Data Included in Analyses by State") %>% kable_styling()
names(countydem)
countydem <- countydem %>% mutate(Total = White + Black + Native.American + Asian + Hispanic, White.pct = White/Total, Black.pct = Black/Total, Native.American.pct = Native.American/Total, Asian.pct = Asian/Total, Hispanic.pct = Hispanic/Total, College.Grad.pct = College.Grad/Total)

countydem$inanalysislabel <- factor(countydem$inanalysis, levels = c(1, 0), labels = c("Counties in States With Abortion Data", "Counties in States Without Abortion Data"))

label(countydem$Women) = "Women"
label(countydem$White.pct) = "White"
label(countydem$Black.pct) = "Black"
label(countydem$Native.American.pct) = "Native American"
label(countydem$Asian.pct) = "Asian"
label(countydem$Hispanic.pct) = "Hispanic"
label(countydem$Median.Income) = "Median Income"
label(countydem$College.Grad.pct) = "College Graduates"
label(countydem$percentdem2008) = "Voting for the Democratic candidate in the most recent presidential election"

units(countydem$Women) = "No."
units(countydem$White.pct) = "%"
units(countydem$Black.pct) = "%"
units(countydem$Native.American.pct) = "%"
units(countydem$Asian.pct) = "%"
units(countydem$Hispanic.pct) = "%"
units(countydem$Median.Income) = "$"
units(countydem$College.Grad.pct) = "%"
units(countydem$percentdem2008) = "%"

# The following function is based on 
# "Using the table1 Package to Create HTML Tables of Descriptive Statistics"
# by Benjamin Rich.  Available at:
# https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html#example-a-column-of-p-values
p_value_t_test <- function(x, ...){
    y <- unlist(x[1:2])
    g <- factor(rep(1:length(x[1:2]), times=sapply(x[1:2], length)))
    p_value <- t.test(y ~ g)$p.value %>% round(3) %>% as.character()
    p_value <- ifelse(p_value == '0', '< 0.001', p_value)
}

table1(~ Women + White.pct + Black.pct + Native.American.pct + Asian.pct + Hispanic.pct + Median.Income + College.Grad.pct + percentdem2008 | inanalysislabel, 
       data = countydem,
       extra.col = list(`P-value` = p_value_t_test),
       footnote = 'P-values are derived by two sample t-tests.'
       )
```


```{r}
#calculate propensity scores
propensitywts <- glm(inanalysis ~ Black.pct + Native.American.pct + Asian.pct + Hispanic.pct + Median.Income + Women + percentdem2008, data=countydem, family=binomial())
summary(propensitywts)

p.score <- predict(propensitywts, countydem, type = "response")
df <- countydem
df$ps <- p.score

#3 plots of distribution of propensity scores 

ggplot(df, aes(x = ps, group = inanalysislabel)) + geom_density(aes(fill = inanalysislabel), alpha = 0.2) +
  xlab('Probability of Being in the Analysis') +
  ggtitle('Propensity Score Distribution by Group') +
  scale_fill_discrete('') +
  theme(legend.position = 'bottom', legend.direction = 'vertical') +
  ylab("Density")


ggplot(df, aes(x = ps, fill = inanalysislabel, color = inanalysislabel)) +
  geom_histogram(alpha = 0.3, position = 'identity', bins=15) +
  facet_grid(as.factor(inanalysis) ~ .) +
  xlab('Probability of Being in the Analysis') +
  ggtitle('Propensity Score Distribution by Group') +
  scale_fill_discrete('') +
  scale_color_discrete('') +
  theme(legend.position = 'bottom', legend.direction = 'vertical')

df %>%
  mutate(ps.grp = round(ps/0.05) * 0.05) %>%
  group_by(inanalysis, ps.grp) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(n2 = ifelse(inanalysis == 0, yes = n, no =  -1*n)) %>%
  ggplot(aes(x = ps.grp, y = n2, fill = as.factor(inanalysis))) +
  geom_bar(stat = 'identity', position = 'identity') +
  geom_text(aes(label = n, x = ps.grp, y = n2 + ifelse(inanalysis == 0, 8, -8))) +
  xlab('Probability of Being in the Analysis') +
  ylab('N') +
  ggtitle('Propensity Score Distribution by Group') +
  scale_fill_discrete(name = "", labels = c("Counties in States With Abortion Data", "Counties in States Without Abortion Data")) +
  scale_x_continuous(breaks = seq(0, 1, 0.05)) +
  theme(legend.position = 'bottom', legend.direction = 'vertical',
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank())

  

```

```{r}
#run models
countydemanalysis <- countydem %>% filter(inanalysis == 1)

model <- lm(AbortionRate2010 ~ restrictionlevel + distancetoprovider + restrictionlevel*distancetoprovider, data = countydemanalysis, weights = ps)
summary(model)

```

```{r}
natality <- read.delim("natalitydata.txt")
maternalmortality <- read.delim("maternalmortality.txt")
infantdeath <- read.delim("infantdeathrecords.txt")

infantdeath$County <- gsub(infantdeath$County, pattern = ", [[:print:]]*$", replacement = "")
infantdeath$County <- gsub(infantdeath$County, pattern = " County", replacement = "")
infantdeath$County <- str_to_title(infantdeath$County)
infantdeath$County <- na_if(infantdeath$County, "")
infantdeath <- infantdeath %>% filter(County != "Unidentified Counties")
infantdeath$County <- gsub(infantdeath$County, pattern = " Parish", replacement = "")
infantdeath$County <- gsub(infantdeath$County, pattern = "Borough", replacement = "Municipality")
infantdeath$County <- gsub(infantdeath$County, pattern = " City", replacement = "")
infantdeath$County[!(infantdeath$County %in% countydem$county)]
infantdeath <- infantdeath %>% rename(state = State, county = County, infantdeathrate = Death.Rate)
infantdeath <- infantdeath %>% dplyr::select(state, county, infantdeathrate)

maternalmortality$County <- gsub(maternalmortality$County, pattern = ", [[:print:]]*$", replacement = "")
maternalmortality$County <- gsub(maternalmortality$County, pattern = " County", replacement = "")
maternalmortality$County <- str_to_title(maternalmortality$County)
maternalmortality$County <- na_if(maternalmortality$County, "")
maternalmortality <- maternalmortality %>% filter(County != "Unidentified Counties")
maternalmortality$County <- gsub(maternalmortality$County, pattern = " Parish", replacement = "")
maternalmortality$County <- gsub(maternalmortality$County, pattern = "Borough", replacement = "Municipality")
maternalmortality$County <- gsub(maternalmortality$County, pattern = " City", replacement = "")
maternalmortality$County <- gsub(maternalmortality$County, pattern = "La Porte", replacement = "Laporte")
maternalmortality$County <- gsub(maternalmortality$County, pattern = "De Kalb", replacement = "Dekalb")
maternalmortality$County <- gsub(maternalmortality$County, pattern = "Mc Kean", replacement = "Mckean")
maternalmortality$County[!(maternalmortality$County %in% countydem$county)] <- gsub(maternalmortality$County[!(maternalmortality$County %in% countydem$county)], pattern = "Municipality", replacement = "Borough")
maternalmortality$County[!(maternalmortality$County %in% countydem$county)] <- gsub(maternalmortality$County[!(maternalmortality$County %in% countydem$county)], pattern = "Borough", replacement = "And Borough")
maternalmortality$County[!(maternalmortality$County %in% countydem$county)] <- gsub(maternalmortality$County[!(maternalmortality$County %in% countydem$county)], pattern = "/", replacement = " ")
maternalmortality$County[!(maternalmortality$County %in% countydem$county)] <- gsub(maternalmortality$County[!(maternalmortality$County %in% countydem$county)], pattern = " Census Area", replacement = "")
maternalmortality$County[!(maternalmortality$County %in% countydem$county)] <- gsub(maternalmortality$County[!(maternalmortality$County %in% countydem$county)], pattern = " And", replacement = "")
maternalmortality$County[!(maternalmortality$County %in% countydem$county)]

maternalmortality <- maternalmortality %>% rename(state = State, county = County, maternaldeaths = Deaths, maternaldeathrate = Crude.Rate)
maternalmortality <- maternalmortality %>% dplyr::select(state, county, maternaldeathrate)

natality$County <- gsub(natality$County, pattern = ", [[:print:]]*$", replacement = "")
natality$County <- gsub(natality$County, pattern = " County", replacement = "")
natality$County <- str_to_title(natality$County)
natality$County <- na_if(natality$County, "")
natality <- natality %>% filter(County != "Unidentified Counties")
natality$County <- gsub(natality$County, pattern = " Parish", replacement = "")
natality$County <- gsub(natality$County, pattern = "Borough", replacement = "Municipality")
natality$County <- gsub(natality$County, pattern = " City", replacement = "")
natality$County <- gsub(natality$County, pattern = "La Porte", replacement = "Laporte")
natality$County[!(natality$County %in% countydem$county)]
natality <- natality %>% rename(state = State, county = County)
natality <- natality %>% dplyr::select(state, county, Birth.Rate, Fertility.Rate)

countydem <- left_join(countydem, infantdeath, by = c("state", "county"))
countydem <- left_join(countydem, maternalmortality, by = c("state", "county"))
countydem <- left_join(countydem, natality, by = c("state", "county"))

sum(is.na(countydem$infantdeathrate)/nrow(countydem))
countydem$maternaldeathrate <- na_if(countydem$maternaldeathrate, "Suppressed")
sum(is.na(countydem$maternaldeathrate)/nrow(countydem))
sum(is.na(countydem$Birth.Rate)/nrow(countydem))

```


\newpage

## Sources 

Data on the county population centroids came from the United States Census Bureau.
Available at https://www.census.gov/geographies/reference-files/time-series/geo/centers-population.html Accessed 12/11/21.

Data on the county-level voting totals came from the Harvard Dataverse.
Available at https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ. Accessed 12/13/21. 


## Code Appendix
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
